Cosmological constraints from the redshift dependence of the volume effect using the galaxy 2-point correlation function across the line-of-sight

### 2017-04-03 draft/draft3.2.tex
updates

\begin{abstract}
We propose to use the redshift dependence of the galaxy 2-point correlation function across the line-of-sight (2pCF), $\xi(r_{\bot})$, to constrain cosmological parameters.
When an incorrect set of cosmological parameters is assumed for transforming galaxy redshifts to comoving distances, 
it produces {\it redshift-dependent mis-scaling} in the galaxy distribution.
As a result, we observe a redshift-dependent rescaling in the measured $\xi(r_{\bot})$.
Using mock galaxy catalogues (containing hundreds of millions of galaxies at redshifts 0, 0.5, 1, 1.5, 2) from the Horizon Run 4 N-body simulation, we find that the redshift evolution of the $\xi(r_{\bot})$ shape, as a function of scale,
is sensitive to the adopted cosmological parameters. 
% mock galaxy catalogues having hundreds of millions of galaxies at redshifts 0, 0.5, 1, 1.5, 2.
However, it is relatively insensitive to the gravitational growth of structure,
the galaxy bias, and the effect of redshift space distortions.
Analyzing the 2pCF shape on scales $5  h^{-1} {\rm Mpc} \leq r_\perp \leq 40 h^{-1} {\rm Mpc}$, 
we derive tight constraints on $\Omega_m$ and $w$ with 68.3\% CL uncertainties of 0.015 and 0.05.
%Our method is complementary to current works using 2pCF to constrain cosmological parameters.
This method could be applicable to future large scale structure surveys, especially photometric surveys such as DES, LSST, 
to derive tight cosmological constraints.
This work is a continuation of our previous works \citep{Li2014,Li2015,Li2016}
as a strategy to constrain cosmological parameters using redshift-invariant physical quantities.
They are all powerful in constraining cosmology and have many merits,
e.g., they are less affected by the systematic effects induced by non-linear clustering or RSD,
they do not require accurate modeling of galaxy clustering,
and are applicable on relatively small clustering scales.
\end{abstract}

@@

The discovery of cosmic acceleration \citep{Riess1998,Perl1999} implies the existence of a ``dark energy'' component in the Universe 
or the breakdown of Einstein's gravity theory on cosmological scales \citep[see ][for a recent review]{2012IJMPD..2130002Y}.
The theoretical explanation and observational probes of cosmic acceleration have attracted 
tremendous attention in the last two decades and are still far from being well understood or accurately measured \citep{SW1989,Li2011,DHW2013}.
%, but its properties still remain far from accurately known and its nature still remains mystery. 
%The simplest and currently favored solution is the additional field called dark energy which has negative pressure and is spatially homogeneous.
%The simplest explanation for dark energy is that it is a cosmological constant or vacuum energy, with constant equation of state (EoS) $w = -1$.

In an effort to probe the cosmic expansion history the large scale structure (LSS) surveys enable measurements of two geometrical quantities, 
the angular diameter distance $D_A$ and the Hubble factor, $H$. 
If they were precisely measured as a function of redshift, 
tight constraints can be placed on cosmological parameters, 
e.g. the matter density $\Omega_m$ and the equation of state (EoS) of dark energy $w$.

When an incorrect cosmological model is assumed when transforming galaxy redshifts to comoving distances,
geometric distortions are produced in the resultant galaxy distribution. It should be noted that this effect is not related to galaxy peculiar velocities which could cause redshift space distortions.
The distortions due to coordinate transforms include the ``volume effect'' which is the mis-scaling of the size of structures,
and the Alcock-Paczynski (AP) effect, the shape distortion induced by the fact that distances along and perpendicular to the line of sight are fundamentally different.
%Measuring the ratio of galaxy clustering in the radial and transverse directions, provides a probe of the Alcock-Paczynski (AP) effect \citep{AP1979}.
%There have been a series of works studying the application of the AP test to the LSS

Various statistical methods have been proposed to measure these effects.
The AP effect was first suggested to be measured through the anisotropic clustering of galaxies \citep{Ballinger1996,Matsubara1996}. Later, with the advent of large observational spectroscopic programs, this was applied to a series of LSS surveys
\citep{Outram2004,Blake2011,ChuangWang2012,Reid2012,Beutler2013,Linder2013,2014arXiv1407.2257S, Jeong2014,Sutter2014,2014ApJ...781...96L,Alam2016, Beutler2016, Sanchez2016}, while considering the detailed effect of systematics for unbiased estimates of $D_A(z)$ and $H(z)$ \citep{2014MNRAS.445....2V,2016MNRAS.tmp.1473R,2016arXiv160302389S}.

However, there have also been alternative approaches in utilising the AP effect using the symmetry properties of galaxy pairs \citep{Marinoni2010,Jennings2011,BB2012}
and cosmic voids \citep{Ryden1995,LavausWandelt1995,Sutter2014,Qingqing2016}.
Methods proposed to measure the volume effect include the number counting of 
galaxy clusters \citep{PS1974,VL1996}, the topology of LSS \citep{topology},
the BAO scale \citep{EHT1998,BG03,SE03},
and the shape of 2pCF and power spectrum \citep{Sanchez2006,Sanchez2009}.

In \cite{Li2014,Li2015,Li2016} we developed a novel strategy to probe $D_A$ and $H$ from the LSS data.
%proposed a series of methods to probe the geometric information encoded in the LSS data.
We found that the AP or volume effect in general introduces a geometric distortion that evolves significantly with redshift;
this phenomenon can be utilized to distinguish the AP effect from the other effects, such as the redshift-space distortion (RSD).
The radial distances of galaxies are inferred from the measured redshifts, which are distorted by the peculiar motion of galaxies, 
leading to apparent distortion in the redshift space galaxy distribution \citep{FOG,Kaiser1987,Ballinger1996}.
It is the major systematic effects limiting our ability to probe the geometry of LSS from the galaxy distribution.
However, we found that the influence of RSD is significantly reduced when we consider only the redshift dependence of the distortion.

\cite{Li2014} attempted to use the {\it galaxy density gradient field} 
to characterize the redshift evolution of the large scale galaxy distribution. 
We tested the idea on the Horizon Run 3 (HR3) N-body simulations \citep{horizonrun},
showing that the redshift evolution of the gradient field 
can be used to infer unbiased constraints on cosmological parameters and that it is, largely, insensitive to RSD.
The same topic was revisited in \cite{Li2015}, 
using a slightly modified version of the usual anisotropic galaxy two-point correlation function (2pCF), $\xi(\mu)$, 
where $\mu\equiv \cos \theta$ and $\theta$ is the angle between the connection of the galaxy pairs and the line of sight direction.
When an incorrect cosmological model is adopted, 
the shape of $\xi(\mu)$ is altered due to the anisotropic distortion of the LSS (the AP effect), 
and the amplitude is shifted due to the change in the comoving volume elements;
both effects have significant redshift dependence.
Although the RSD effect significantly distorts $\xi(\mu)$, 
it exhibits much less redshift evolution.
The method was applied to BOSS DR12 galaxies \citep{Li2016} to probe the redshift dependence of the AP effect, 
and we arrive at tight cosmological constraints competitive with the mainstream probes of SNIa, BAO, and CMB.

In this paper we continue our exploration of this strategy and present 
a new method using the redshift evolution of the shape of the galaxy 2pCF across the line-of-sight as a function of radial scale, $\xi(r_{\bot})$,
to constrain cosmology.
A wrongly adopted cosmology results in {\it redshift-dependent mis-scaling} of the galaxy distribution,
which leads to redshift-evolution of the shape of $\xi(r_{\bot})$.
The signal is sensitive to the cosmological parameters but relatively insensitive to the other effects such as the gravitational growth of structure,
the galaxy bias, and the RSD effect.
We apply our method to HR4 mock catalogues, using hundreds of millions of galaxies at redshifts $0\leq z\leq2$
and develop the statistical procedure to constrain cosmological parameters.

%In this paper we continue our series of work and probe to use the angular correlation function as a statistical tool to probing the volume effect in the angular direction.
%The method of redshift dependent AP effect has been shown great power to constrain cosmological parameters \cite{Li2014,Li2015,Li2015}.
%We hope this method also.
%The two methods probe different information encoded in the 2pCF and can be combined to fully probe the geometrical effect in LSS.
%Or one can also use the redshift evolution of 2D 2pCF xi(s,mu) to have all information included.
%We developed a method measuring the redshift of the shape of galaxy angular correlation function, $\hat \xi_{r_\perp}$, to constrain cosmology.
%A wrongly adopted cosmology result in redshift-dependent mis-scaling in the constructed galaxy distribution,
%which shifts apparent scale of the clustering patterns;
%this manifests itself as a redshift-dependent rescaling of the measured angular correlation function.
%We find this signal is sensitive to the adopted cosmological parameters and relatively insensitive to the gravitational growth of structure,
%the galaxy bias, and the RSD effect.
%We test our method on the HR4 mock galaxies.
%The method can be applied to current and future LSS surveys,
%especially, the photometric surveys such as DES and LSST which probe a large amount of galaxies and the large redshift error will not quite affect the angular 2pCF measurement.

The outline of this paper proceeds as follows. 
In \S 2 we briefly review the nature and consequences of the LSS geometric distortion when performing coordinate transforms in a cosmological context. 
In \S 3 we describe the N-body mock galaxies used to test our method.
The methodology is presented in \S 4.
We conclude in \S 5.

@@

%As will be introduced in the next section, 
%The output of HR4 simulation includes one all-sky light cone mock galaxy catalogue reaching $r=3\,150$ ${h^{-1}}$Mpc 
%and a series ({\bf how many?}) of snapshot mock galaxy catalogues at different redshifts.
We take the snapshot data of the mock galaxies at $z=0,0.5,1,1.5,2$.
%We require subhalos to have at least 30 member particles, 
%so the minimal mass of galaxies is $\simeq m_{p} \simeq 2.7 \times 10^{11} \hMsun$.
Setting a minimal halo mass of $3\times 10^{11} \hMsun$, 
%$4.57\times 10^8$, $4.06\times 10^8$, $3.52\times 10^8$, $3.06\times 10^8$ and $2.28 \times 10^8$ mock galaxies,
we select 457, 406, 352, 306 and 228 million mock galaxies at the five redshifts,
corresponding to a number density of 
%$1.46\times 10^{-2}$, $1.30\times 10^{-2}$, $1.13\times 10^{-2}$, $0.98\times 10^{-2}$ and $0.73\times 10^{-2}\ (h^{-1} \rm Mpc)^{-3}$,
1.46, 1.30, 1.13, 0.98 and 0.73 in units of $ 10^{-2} h^{3} \rm Mpc^{-3}$
%$1.46\times 10^{-2}$, $1.30\times 10^{-2}$, $1.13\times 10^{-2}$, $0.98\times 10^{-2}$ and $0.73\times 10^{-2}\ (h^{-1} \rm Mpc)^{-3}$,
respectively.
Applying a uniform mass cut, at higher redshift we obtained a smaller number of galaxies with {\it larger bias}.

@@

The predicted constraints from this method clearly wins due to the large sample size.
The total number of galaxies used in this work is 1,749 million,
which is 1,000 times of the BOSS DR12 sample size.
So the constrained area is also $\sim 30$ times smaller.
Such a large sample size will be enabled by LSST (the Large Synoptic Survey Telescope) \footnote{https://www.lsst.org/}, 
which probe tens of billions of galaxies at $z\lesssim2$.

>>>>

\begin{abstract}
We develop a methodology to use the redshift dependence of the galaxy 2-point correlation function across the line-of-sight (2pCF), $\xi(r_{\bot})$, as a probe of cosmological parameters. 
The positions of galaxies in comoving Cartesian space varies under different cosmological parameter choices, 
inducing a {\it redshift-dependent scaling} in the galaxy distribution. 
This geometrical distortion can be observed as a redshift-dependent rescaling in the measured $\xi(r_{\bot})$. 
We test this methodology using a sample of 1.75 billion mock galaxies at redshifts 0, 0.5, 1, 1.5, 2, 
drawn from the Horizon Run 4 N-body simulation.
The shape of $\xi(r_{\bot})$ can exhibit a significant redshift evolution when the galaxy sample is 
analyzed under a cosmology differing from the true, simulated one. 
Other contributions, including the gravitational growth of structure, 
galaxy bias, and the redshift space distortions, 
do not produce large redshift evolution in the shape. 
We show that one can make use of this geometrical distortion to constrain the values of 
cosmological parameters governing the expansion history of the universe. 
This method could be applicable to future large scale structure surveys, 
especially photometric surveys such as DES, LSST, to derive tight cosmological constraints. 
This work is a continuation of our previous works \citep{Li2014,Li2015,Li2016} 
as a strategy to constrain cosmological parameters using redshift-invariant physical quantities.
%We propose to use the redshift dependence of the galaxy 2-point correlation function across the line-of-sight (2pCF), $\xi(r_{\bot})$, to constrain cosmological parameters.
%When an incorrect set of cosmological parameters is assumed for transforming galaxy redshifts to comoving distances, 
%it produces {\it redshift-dependent mis-scaling} in the galaxy distribution.
%As a result, we observe a redshift-dependent rescaling in the measured $\xi(r_{\bot})$.
%We test this idea using a sample of 1.75 billion mock galaxies at redshifts 0, 0.5, 1, 1.5, 2,
%drawn from the Horizon Run 4 N-body simulation.
%The shape of $\xi(r_{\bot})$ exhibits redshift evolution when the adopted cosmology is wrong,
% mock galaxy catalogues having hundreds of millions of galaxies at redshifts 0, 0.5, 1, 1.5, 2.
%while the other effects, including the gravitational growth of structure,
%galaxy bias, and the redshift space distortions, do not produce large redshift evolution in the shape.
%Analyzing the 2pCF shape on scales $20  h^{-1} {\rm Mpc} \leq r_\perp \leq 40 h^{-1} {\rm Mpc}$, 
%we derive tight constraints on $\Omega_m$ and $w$ with 68.3\% CL uncertainties of 0.015 and 0.05.
%We show that one can make use of this fact to constrain the values of cosmological parameters governing the 
%expansion history of the universe.
%Our method is complementary to current works using 2pCF to constrain cosmological parameters.
%This method could be applicable to future large scale structure surveys, 
%especially photometric surveys such as DES, LSST, 
%to derive tight cosmological constraints.
%This work is a continuation of our previous works \citep{Li2014,Li2015,Li2016}
%as a strategy to constrain cosmological parameters using redshift-invariant physical quantities.
%They are all powerful in constraining cosmology and have many merits,
%e.g., they are less affected by the systematic effects induced by non-linear clustering or RSD,
%they do not require accurate modeling of galaxy clustering,
%and are applicable on relatively small clustering scales.
\end{abstract}

@@

The discovery of cosmic acceleration \citep{Riess1998,Perl1999} implies either the existence of a ``dark energy'' component in the Universe 
or the breakdown of Einstein's gravity theory on cosmological scales \citep[see ][for a recent review]{2012IJMPD..2130002Y}.
The theoretical explanation and observational probes of cosmic acceleration have attracted 
tremendous attention in the last two decades and are still far from being well understood or accurately measured \citep{SW1989,Li2011,DHW2013}.
%, but its properties still remain far from accurately known and its nature still remains mystery. 
%The simplest and currently favored solution is the additional field called dark energy which has negative pressure and is spatially homogeneous.
%The simplest explanation for dark energy is that it is a cosmological constant or vacuum energy, with constant equation of state (EoS) $w = -1$.

In an effort to probe the cosmic expansion history, large scale structure (LSS) surveys are utilized to  measure two key geometrical quantities; 
the angular diameter distance $D_A$ and the Hubble factor, $H$. 
If they were precisely measured as a function of redshift, 
then tight constraints can be placed on cosmological parameters, 
e.g. the matter density $\Omega_m$ and the equation of state (EoS) of dark energy $w$.

When an incorrect cosmological model is assumed for transforming galaxy redshifts to comoving distances,
geometric distortions are induced in the resultant galaxy distribution. 
It should be noted that this effect is not related to the peculiar velocities of galaxies, which cause redshift-space distortions (RSD).
The distortions due to coordinate transforms include the ``volume effect'' which is the mis-scaling of the size of structures, 
and the Alcock-Paczynski (AP) effect, the shape distortion induced by the fact that distances along and perpendicular to the line of sight are fundamentally different.

Various statistical methods have been proposed to measure these effects.
The AP effect was first suggested to be measured through the anisotropic clustering of galaxies \citep{Ballinger1996,Matsubara1996}. 
Later, with the advent of large observational spectroscopic programs, this was applied to a series of LSS surveys
\citep{Outram2004,Blake2011,ChuangWang2012,Reid2012,Beutler2013,Linder2013,2014arXiv1407.2257S, Jeong2014,Sutter2014,2014ApJ...781...96L,Alam2016, Beutler2016, Sanchez2016}, 
while considering the detailed effect of systematics for unbiased estimates of $D_A(z)$ and $H(z)$ \citep{2014MNRAS.445....2V,2016MNRAS.tmp.1473R,2016arXiv160302389S}.
Alternative approaches include utilizing the AP effect using the symmetry properties of galaxy pairs \citep{Marinoni2010,Jennings2011,BB2012}
and cosmic voids \citep{Ryden1995,LavausWandelt1995,Sutter2014,Qingqing2016}.

There have also been studies proposing to measure the volume effect of this geometrical rescaling, 
these include the number counting of galaxy clusters \citep{PS1974,VL1996}, 
the topology of LSS \citep{topology}, the BAO scale \citep{EHT1998,BG03,SE03}, 
and the shape of 2pCF and power spectrum \citep{Sanchez2006,Sanchez2009}.

In \cite{Li2014,Li2015} we developed a novel strategy to probe $D_A$ and $H$ from the LSS data. 
We found that the AP effect introduces a geometric distortion that evolves significantly with redshift;
this phenomenon can be utilized to distinguish the AP effect from the effect of RSD.

The radial distances of galaxies are inferred from their measured redshifts, 
which are distorted due to the peculiar motion of galaxies. 
This leads to an apparent distortion in the redshift-space galaxy distribution \citep{FOG,Kaiser1987,Ballinger1996}. 
This RSD effect is the main systematic limiting our ability to probe the geometry of LSS from the galaxy distribution. 
However, we found that the influence of RSD is significantly reduced when focusing on the redshift dependence of galaxy clustering.

In \cite{Li2016}, we analyzed the two-point correlation function (2pCF) of galaxies measured in the Baryon Oscillation Spectroscopic Survey (BOSS). 
We showed that the redshift dependence of the AP effect leads to tight cosmological constraints that are competitive with mainstream cosmological probes such as type Ia supernovae (SNIa), baryon acoustic oscillations (BAO), and cosmic microwave background (CMB).

In this paper we continue our exploration of redshift invariant galaxy clustering 
and present a new method for constraining cosmology using the redshift evolution of 
the galaxy two-point correlation function (2pCF) across the line-of-sight, $\xi(r_{\bot})$. 
An incorrect choice of cosmology results in {\it redshift-dependent scaling} of the galaxy distribution. 
This, in turn, results in a significant redshift evolution in the shape of $\xi(r_{\bot})$.
Of course the correlation function has other contributions factors that may induce redshift evolution, 
such as the gravitational growth of structure, galaxy bias, and redshift space distortions. 
However we find that these other contributions do not produce large redshift evolution in the shape of $\xi(r_{\bot})$. 
We test this methodology using a sample of 1.75 billion mock galaxies at redshifts 0, 0.5, 1, 1.5, 2, drawn from the Horizon Run (HR) 4 N-body simulation.

The outline of this paper proceeds as follows. 
In \S 2 we briefly review the nature and consequences of the LSS geometric distortion when performing coordinate transforms in a cosmological context. 
In \S 3 we describe the N-body mock galaxies used in the analysis.
The methodology is presented in \S 4 and \S 5.
We conclude in \S 6.

@@

%As will be introduced in the next section, 
%The output of HR4 simulation includes one all-sky light cone mock galaxy catalogue reaching $r=3\,150$ ${h^{-1}}$Mpc 
%and a series ({\bf how many?}) of snapshot mock galaxy catalogues at different redshifts.
We take the snapshot data of the mock galaxies at $z=0,0.5,1,1.5,2$.
%We require subhalos to have at least 30 member particles, 
%so the minimal mass of galaxies is $\simeq m_{p} \simeq 2.7 \times 10^{11} \hMsun$.
Setting a minimal halo mass of $3\times 10^{11} \hMsun$, 
%$4.57\times 10^8$, $4.06\times 10^8$, $3.52\times 10^8$, $3.06\times 10^8$ and $2.28 \times 10^8$ mock galaxies,
we select 457, 406, 352, 306 and 228 million mock galaxies at the five redshifts,
corresponding to a number density of 
%$1.46\times 10^{-2}$, $1.30\times 10^{-2}$, $1.13\times 10^{-2}$, $0.98\times 10^{-2}$ and $0.73\times 10^{-2}\ (h^{-1} \rm Mpc)^{-3}$,
1.46, 1.30, 1.13, 0.98 and 0.73 in units of $ 10^{-2} h^{3} \rm Mpc^{-3}$
%$1.46\times 10^{-2}$, $1.30\times 10^{-2}$, $1.13\times 10^{-2}$, $0.98\times 10^{-2}$ and $0.73\times 10^{-2}\ (h^{-1} \rm Mpc)^{-3}$,
respectively.
The total number of galaxies is 1.75 billion.
Applying a uniform mass cut, at higher redshift we obtained a smaller number of galaxies with {\it larger bias}.

@@

The predicted constraints from this method clearly wins due to the large sample size.
The total number of galaxies used in this work is 1.75 billion,
which is 1,000 times of the BOSS DR12 sample size.
So the constrained area is also $\sim 30$ times smaller.
Such a large sample size will be enabled by LSST (the Large Synoptic Survey Telescope) \footnote{https://www.lsst.org/}, 
which probe tens of billions of galaxies at $z\lesssim2$.
